Field and experimental data together with computational models

### 2022-02-09 manuscript_1.0/manuscript_1.0.tex
Changes addressing Zegni's comment in the ms

% For Pandoc highlighting needs

% Pandoc citation processing

@@

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
While it is generally straightforward to quantify individual performance in cognitive experiments, identifying the underlying cognitive processes remains a major challenge. Often, different mechanistic underpinnings may yield similar performances, and Lloyd Morgan's cannon warrants acceptance of the simpler explanation. Alternatively, when the different mechanisms interact with environmental conditions, variation in performance across environments might allow to statistically infer the mechanism responsible. However, to do this it is necessary to have quantitative predictions for the different candidate mechanisms. Here, we use a set of computational models to get quantitative predictions from alternative learning mechanisms, and using Bayesian statistics fit the model parameters to performance data. We used experimental data on performance in an ephemeral reward task by wild-caught cleaner fish \emph{Labroides dimidiatus}, as well as cleaner and client fish densities from the locations of capture. The task can in principle be solved by estimating future consequences of an action, or by perceiving the removal of the ephemeral reward as psychological punishment (negative reinforcement). We found that a model where cleaners only estimate the future consequences of their actions explains best which cleaner relative abundances cause the fish to develop a preference for an ephemeral food source. This model also yields performances that can be considered the result of locally optimal decision-rules, in contrast to the negative reinforcement model. We argue that the combination of computational models with data is a powerful tool to infer the mechanistic underpinning behind cognitive performance.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Some pieces required from the pandoc template
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\providecommand{\EndFirstPage}{%
}

@@

Cooperation among individuals of the same or different species is
ubiquitous in nature. Whenever such interactions involve some degree
of partner choice, biological market theory predicts that the exchange
rate of goods and/or services traded adjusts to the law of supply and demand
\citep{noe_Biological_1995a}. While biological market models have largely focused
on evolutionary time scales \citep{noe_Biological_1994a, johnstone_Mutualism_2008a}
\(\color{red}{\text{Hoekstra xxxx}}\), supply and demand conditions, which typically
depend on the abundance of the species involved, certainly vary in time
and space. Therefore, natural selection should
favour the ability to flexibly adjust decisions and behavioural output to
current market conditions. Indeed, such adjustments have been repeatedly
documented \citep{axen_Signalling_1996}. In animals, an obvious general candidate
mechanism for the strategic adjustment is the cognitive machinery. However,
it is not clear which cognitive mechanisms allow individuals to adjust their
behaviour to the varying market conditions. As adjustments to changes in
market conditions have been documented in a great variety of species with
highly variable brain anatomies, the question arises to what extent mechanisms
beyond basic associative learning may be involved.

One example of strategic adjustment in a biological market is the marine
cleaning mutualism involving the cleaner fish \emph{Labroides dimidiatus} and
`client' fish. Clients offer themselves as food patches for cleaners to
remove ectoparasites, which gives the cleaners calories and the clients
improved health \citep{waldie_LongTerm_2011, ros_Does_2011, triki_Effects_2016}. As cleaners have about 2000
interactions per day \citep{grutter_Relationship_1995}, they regularly face
the situation that two clients seek cleaning simultaneously. In such
conditions, it is important to distinguish between `resident' and
`visitor' client species \citep{bshary_Cleaner_2001a}. The residents have
small home ranges and hence access to only the local cleaner, while the
visitors have large home ranges and hence access to two or more
cleaners. Thus, visitors have the option to switch to another cleaner if
being made to wait, while residents must wait for inspection. Indeed,
visitors have been observed to use their partner choice options in that
way \citep{bshary_Choosy_2002}, which may explain why cleaners give
visitors service priority in a field study in the Red Sea
\citep{bshary_Cleaner_2001a}. More generally, cleaners apparently prefer to
interact with visitors \citep{triki_Marine_2021}. Such a preference may be
linked to a client's value as a food patch; visitors are on average
larger than residents \citep{bshary_Cleaner_2001a}, and size is positively
correlated with parasite load \citep{grutter_Relationship_1995}. However,
early laboratory experiments showed that leaving causes service
priority independent of client value \citep{bshary_Asymmetric_2002}.
The experiments consisted of asimultaneous two-choice task,
where cleaners had to decide which of two
Plexiglas plates, which offered equal amounts of food, to approach first
and eat mashed prawn of it. One plate was made to behave like a
resident, i.e.~it would remain until the cleaner had eaten off it. The
other plate was made to behave like a visitor, i.e.~it would be removed
if not inspected first. Cleaners learned to prefer the visitor plate and
hence obtained the double amount of food \citep{bshary_Asymmetric_2002}.
Adult cleaner fish outperform various primates as well as rats and
pigeons in this original version of the biological market task
\citep{salwiczek_Adult_2012, zentall_Early_2017}; only African grey
parrots solve this task as well \citep{pepperberg_Can_2014}. As the correct
solution must be learned, and primates are known to readily distinguish
between one and two reward units if presented in more conventional ways
\(\color{red}{\text{ref}}\), what makes the task difficult for most non-cleaners?

Recent research on the cognitive tool kit needed to solve an ephemeral
reward task has used a broad approach as proposed by Shettleworth
\citep{shettleworth_Cognition_2009},
who defined cognition as including all ways in which animals take in information
through the senses, process, retain and decide to act on it. Indeed, perception
of relevant cues is of major importance for individual performance;
i.e.~what relevant information is encoded in the design of the two
plates (colour versus size; \citep{wismer_Cuebased_2019}), or whether
the information is in the colour of
the food \citep{pretot_Comparative_2021, pretot_Comparing_2016}, and
whether the food is visible or hidden/released only when the subjects
made choices on a touch screen \citep{pretot_Factors_2016}. However, identifying
cues as salient is not the only challenge of the ephemeral reward task,
as revealed by proximate learning models. Such models allow varying the
cognitive tool kit and to evaluate which minimal kit is necessary to solve
the task at hand (e.g. \citep{dubois_Model_2021}). Applied to the ephemeral
reward task, learning models showed that basic reinforcement learning does
not suffice to solve the task \citep{prat_Modelling_2022, quinones_Reinforcement_2019}.
This is particularly so when models assume the more complex natural
situation in which cleaners not only face resident-visitor pairs but also
visitor-visitor and resident-resident pairs, as well only a resident or
only a visitor. To be able to give visitors priority over residents,
cleaners need to be able to assess a client's value separately for the
three possible forms of encounter (alone, paired with a fish with the
same strategic option, paired with a fish with the alternative strategic option)
\citep{quinones_Reinforcement_2019}. The ability to keep such separate scores
for the same stimulus has been termed configurational learning,
or chunking, or segmentation (see references in \citep{prat_Modelling_2022}).

To solve the ephemeral reward task, cleaners also need to account for
the future consequences of current decisions. In the model by
Quiñones \emph{et al.} \citep{quinones_Reinforcement_2019}, this could be achieved in
two non-mutually exclusive ways: through low temporal discounting of
future effects, also termed `chaining' \citep{enquist_Power_2016}; and/or
through perceiving a visitor client leaving as psychological punishment
(i.e.~as a negative reinforcer). Given that a negative reinforcer is part
of the ubiquitous learning mechanism termed operant conditioning
\citep{thorndike_Animal_1898, skinner_Behavior_1938}, this explanation
could be considered the simpler one. Furthermore, accounting for the
future effects of an action is considered an advanced cognitive
ability \citep{suddendorf_Evolution_2007}. Thus, standard logic of
Lloyd Morgan's cannon demands that operant conditioning as the simpler
explanation is to be accepted by default. Here, we explore the
possibility that fitting a model in which the action of these two
mechanisms, operant conditioning and chaining, depends
on model parameters may reveal which process cleaners actually use.
This is because one process may be better at predicting the observed
variation in cleaner fish performance as a function of ecological conditions.

Various repetitions of the ephemeral reward task in different years and
with cleaners caught from various locations around Lizard Island,
Great Barrier Reef, revealed that cleaner and client densities in
nature affect how cleaners perform in the original market task:
when relative densities are low, cleaners are more likely to fail to
complete the task \citep{triki_Biological_2019, triki_Decrease_2018, wismer_Variation_2014}. The explanation for this pattern
is twofold. First, low cleaner abundances implies low supply and high demand
of cleaning services. Under this conditions, there will be less occasions under
which failing to give priority to a visitor translates to an empty
cleaning station. Secondly and also related to the state of the cleaning market,
visitor clients are less likely to exert their choice options
(and hence wait for inspection if made to wait) when
cleaner fish densities are low \citep{triki_Brain_2020}. If visitors are not choosy,
cleaners should not give them priority anymore. Thus, the underling
cognitive machinery and ecological variables apparently interact to produce
such variable performance. By fitting our model to the empirical data on
fish densities and cleaner fish performance in the ephemeral reward task
we aimed to address two questions. First, we aimed to determine the likely
mechanism used by cleaners to incorporate future consequences of current
decisions by testing whether operant conditioning, chaining, or a combination
of both best explains performance. Second and related to the first
question, we aimed to determine whether the two mechanisms differed
with respect to the ecological conditions that are likely to cause high
versus low performance in the ephemeral reward task. If that were the case,
we would assess which mechanism would yield more appropriate performance
patterns. Relying on the logic of biological market theory, we predicted
that appropriate performance is to show low preference for visitors under high
local client-to-cleaner ratio as visitor leaving rates should be low
and any leaving visitor will be readily replaced by another client. In contrast,
cleaners should show high performance if the local client-to-cleaner
ratio favours clients.

>>>>

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

@@

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
While it is generally straightforward to quantify individual performance in cognitive experiments, identifying the underlying cognitive processes remains a major challenge. Often, different mechanistic underpinnings yield similar performances, and Lloyd Morgan's cannon warrants acceptance of the simpler explanation. Alternatively, when the different mechanisms interact with environmental conditions, variation in performance across environments might allow to statistically infer the mechanism responsible. However, to do this it is necessary to have quantitative predictions for the different candidate mechanisms. Here, we use a set of computational models to get quantitative predictions from alternative learning mechanisms, and using Bayesian statistics fit the model parameters to performance data. We used experimental data on performance in an ephemeral reward task by wild-caught cleaner fish \emph{Labroides dimidiatus}, as well as cleaner and client fish densities from the locations of capture. The task can in principle be solved by estimating future consequences of an action, or by perceiving the removal of the ephemeral reward as psychological punishment (negative reinforcement). We found that a model where cleaners only estimate the future consequences of their actions explains best which cleaner relative abundances cause the fish to develop a preference for an ephemeral food source. This model also yields performances that can be considered the result of locally optimal decision-rules, in contrast to the negative reinforcement model. We argue that the combination of computational models with data is a powerful tool to infer the mechanistic underpinning behind cognitive performance.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

\providecommand{\EndFirstPage}{%
}

@@

Disentangling the mechanistic underpinnings of behaviour
is typically done under controlled conditions in a laboratory setting. However,
controlled laboratory conditions often mask the inter and intra-specific
variation that arises in the interaction between mechanisms and
environmental variation. From a evolutionary perspective, mechanisms
are likely selected because of how they allow individuals to respond to
environmental variation. For example, biological market theory predicts that
the exchange rate of goods and/or services traded between cooperative
partners adjusts to the law of supply and demand, when individuals have
some degree of partner choice \citep{noe_Biological_1995a}.
Supply and demand conditions, which typically
depend on the abundance of the species involved, certainly vary in time
and space. Therefore, natural selection should
favour the ability to flexibly adjust decisions and behavioural output to
current market conditions. Indeed, such adjustments have been repeatedly
documented \citep{axen_Signalling_1996}. In animals, an obvious general candidate
mechanism for the strategic adjustment is the cognitive machinery. However,
it is not clear which cognitive mechanisms allow individuals to adjust their
behaviour to the varying market conditions. As adjustments to changes in
market conditions have been documented in a great variety of species with
highly variable brain anatomies, the question arises to what extent mechanisms
beyond basic associative learning may be involved.

One example of strategic adjustment in a biological market is the marine
cleaning mutualism involving the cleaner fish \emph{Labroides dimidiatus} and
`client' fish. Client fish seek cleaner fish services at their territory
(so-called ``cleaning station'') and offer themselves as food patches
to get their ectoparasites removed, which provides cleaners
with food and clients with improved health \citep{waldie_LongTerm_2011, ros_Does_2011, triki_Effects_2016, demaire_Reduced_2020}.
Given the capacity of some client fish to swim larger distances and
access multiple cleaning stations while others access the only cleaning
station at their territory, it is crucial to categorize clients as
either ``visitors'' or ``residents'', respectively. During cleaning interactions,
a cleaner fish often face a choice between a visitor and a resident client
seeking its cleaning services simultaneously. Visitors have the option
to switch to another cleaner fish if being made to wait, while residents
must wait for inspection. Indeed,
visitors have been observed to use their partner choice option in that
way \citep{bshary_Choosy_2002}, which may explain why cleaners give
visitors service priority in a field study in the Red Sea
\citep{bshary_Cleaner_2001a}.

When researchers aimed at testing wild-caught cleaner fish, as
well as individuals from other species, abilities
to prioritize visitor over resident clients in a lab-based paradigm,
they used Plexiglas plates of different colours and shapes offering same
amount of food as surrogates for visitor and resident clients.
One plate is made to behave like a resident, i.e.~it remains until
the cleaner had eaten off it. The other plate was made to behave
like a visitor, i.e.~it is removed if not inspected first.
Cleaners learned to prefer the visitor plate and
hence obtained the double amount of food \citep{bshary_Asymmetric_2002}.
Furthermore, adult cleaner fish outperformed various primates as well as rats and
pigeons in this original version of the biological market task
\citep{salwiczek_Adult_2012, zentall_Early_2017}; yet the African grey
parrots solve this task as well \citep{pepperberg_Can_2014}. Given that primates
are known to readily distinguish
between one and two reward units if presented in more conventional ways
\(\color{red}{\text{ref}}\), this raises the question:
what makes the task difficult for most non-cleaners?

Recent research on the cognitive tool kit needed to solve the ephemeral
reward task has used a broad approach as proposed by Shettleworth
\citep{shettleworth_Cognition_2009}, who defined cognition as including
all ways in which animals acquire information
through the senses, process, retain and decide to act on it.
Indeed, perception of relevant cues is of major importance
for individual performance. For example, some animals find relevant
information in the plates
\citep{wismer_Cuebased_2019}) while others find it in the food
\citep{pretot_Comparative_2021, pretot_Comparing_2016}. However, identifying
cues as salient is not the only challenge of the ephemeral reward task,
as revealed by proximate learning models. Such models allow varying the
cognitive tool kit and to evaluate which minimal kit is necessary to solve
the task at hand (e.g. \citep{dubois_Model_2021}). Applied to the ephemeral
reward task, learning models showed that basic reinforcement learning does
not suffice to solve the ephemeral reward task \citep{prat_Modelling_2022, quinones_Reinforcement_2019}.
This is particularly so when models assume the more complex natural
situation in which cleaner fish not only face resident-visitor pairs but also
visitor-visitor and resident-resident pairs, as well a resident alone or
a visitor alone. To be able to give visitors priority over residents,
cleaners need to be able to assess a client's value separately for the
three possible scenarios (alone, paired with a fish with the
same strategic option, paired with a fish with the alternative strategic option)
\citep{quinones_Reinforcement_2019}. The ability to distinguish and value differently
one stimulus alone from compound versions of it has been termed
configurational learning, or chunking,
or segmentation (see references in \citep{prat_Modelling_2022}).

To solve the ephemeral reward task, cleaners need to account for
the future consequences of current decisions. In the model by
Quiñones \emph{et al.} \citep{quinones_Reinforcement_2019}, this could be achieved in
two non-mutually exclusive ways: through low temporal discounting of
future effects, also termed `chaining' \citep{enquist_Power_2016}; and/or
through perceiving a visitor client leaving as psychological punishment
(i.e.~as a negative reinforcer). Low temporal discounting is when
individuals include in their valuation of an action the reward effects
that this will have in the future. This is done by combining in a single
valuation the reward obtained in the current time with all the reward that
comes after, discounting for how far in the future the future reward
is accrued. `Chaining' the reward of these different time steps allows
individuals to take actions that increase the long-term reward at the sacrifice
of short term considerations \citep{enquist_Power_2016}. Even though, `chaining'
can be readily implemented computationally in learning models
\citep{enquist_Power_2016, sutton_Reinforcement_2018}, cognitively it seems to be
a complex adaptation \citep{suddendorf_Evolution_2007}. On the other hand,
using client behaviour as a negative reinforcer is in principle
easier to implement. Negative reinforcement is part of the ubiquitous
learning mechanism termed operant conditioning \citep{thorndike_Animal_1898, skinner_Behavior_1938}. Thus, standard logic of
Lloyd Morgan's cannon demands that operant conditioning as the simpler
explanation is to be accepted by default. Ideally, the two mechanisms should
be evaluated in light of how well they explain the available data.

Over the last decade, our team tested over a hundred of wild-caught
cleaner fish in the exact same paradigm of of the ephemeral reward
task \citep{salwiczek_Adult_2012, wismer_Variation_2014, triki_Decrease_2018, triki_Biological_2019, triki_Brain_2020}. These fish often come from
different reef locations around the study field site at Lizard Island,
Great Barrier Reef in Australia. Further investigation of the local
eco-sociological conditions revealed that cleaner and client
fish population densities have a substantial impact on cleaner fish
performance in the task. Cleaner fish from reef sites with relatively
low densities were more likely to fail solving the task \citep{triki_Biological_2019, triki_Decrease_2018, wismer_Variation_2014}. The explanation for this pattern
is twofold. First, low cleaner fish population densities imply
low supply and high demand for cleaning services.
Under such conditions, there are fewer occasions under
which failing to give priority to a visitor translates to an empty
cleaning station. Second and also related to the state of the cleaning market,
visitor clients at low cleaner fish population densities
are less likely to exert their partner choice options
and are hence more likely to wait for inspection if not given cleaning
priority \citep{triki_Brain_2020}. In the absence of partner choice
cleaners should not give priority to visitors any more.

The interplay between the underling cognitive machinery and
local ecological conditions apparently generate the documented variable
performance among individuals of the same species.
Our approach of fitting the computational model to the empirical data on
fish densities and cleaner fish censuses and cleaner fish performance
in the ephemeral reward task aimed at: (i), determining which
mechanism cleaner fish use to incorporate future consequences of current
decisions by testing whether operant conditioning, chaining, or a combination
of both best explains their performance; (ii) determining whether the
two mechanisms differed with respect to the ecological conditions
that are likely to cause high versus low performance in the
ephemeral reward task. Additionally, we assessed which mechanism
yields optimal performance patterns. Relying on the logic of
biological market theory, we predicted
that appropriate performance is to show low preference for visitors under low
local cleaner-to-client ratio. In this case, as visitor leaving rates
should be low as well and any leaving visitor will be readily
replaced by another client in the queue. In contrast,
cleaners should show high performance in the ephemeral reward task
if the local cleaner-to-client ratio favours clients.

### 2022-02-09 manuscript_1.0/manuscript_word.Rmd
Changes addressing Zegni's comment in the ms

---
title: Combination of field and experimental data together with computational models
  reveal cognitive mechanisms behind cleaner fish behaviour
author:
- Andrés E. Quiñones^[Institute of Biology, University of Neuchâtel, Neuchâtel, Switzerland
  - andresqp@gmail.com]
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
  pdf_document: default
fig_caption: yes
fontsize: 12pt
header-includes:
- \usepackage{float}
- \usepackage{setspace}\doublespacing
- \linespread{2}
- \usepackage{lineno}
- \linenumbers
- \newcommand{\beginsupplement}{ \setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
keep_tex: yes
linespread: 2
bibliography: Cleanerlearning.bib
style: M:\Zotero\styles\behavioral-ecology-and-sociobiology.csl
spacing: double
abstract: |
  While it is generally straightforward to quantify individual performance in cognitive experiments, identifying the underlying cognitive processes remains a major challenge. Often, different mechanistic underpinnings may yield similar performances, and Lloyd Morgan’s cannon warrants acceptance of the simpler explanation. Alternatively, when the different mechanisms interact with environmental conditions, variation in performance across environments might allow to statistically infer the mechanism responsible. However, to do this it is necessary to have quantitative predictions for the different candidate mechanisms. Here, we use a set of computational models to get quantitative predictions from alternative learning mechanisms, and using Bayesian statistics fit the model parameters to performance data. We used experimental data on performance in an ephemeral reward task by wild-caught cleaner fish *Labroides dimidiatus*, as well as cleaner and client fish densities from the locations of capture. The task can in principle be solved by estimating future consequences of an action, or by perceiving the removal of the ephemeral reward as psychological punishment (negative reinforcement). We found that a model where cleaners only estimate the future consequences of their actions explains best which cleaner relative abundances cause the fish to develop a preference for an ephemeral food source. This model also yields performances that can be considered the result of locally optimal decision-rules, in contrast to the negative reinforcement model.  We argue that the combination of computational models with data is a powerful tool to infer the mechanistic underpinning behind cognitive performance. 
keywords: learning, behaviour, cleaners, bayesian statisitics
---

>>>>

---
title: Field and experimental data together with computational models
  reveal key cognitive mechanisms behind cleaner fish behaviour
author:
- Andrés E. Quiñones^[Institute of Biology, University of Neuchâtel, Neuchâtel, Switzerland
  - andresqp@gmail.com]
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
  pdf_document: default
fig_caption: yes
fontsize: 12pt
header-includes:
- \usepackage{float}
- \usepackage{setspace}\doublespacing
- \linespread{2}
- \usepackage{lineno}
- \linenumbers
- \newcommand{\beginsupplement}{ \setcounter{table}{0} \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
keep_tex: yes
linespread: 2
bibliography: Cleanerlearning.bib
style: M:\Zotero\styles\behavioral-ecology-and-sociobiology.csl
spacing: double
abstract: |
  While it is generally straightforward to quantify individual performance in cognitive experiments, identifying the underlying cognitive processes remains a major challenge. Often, different mechanistic underpinnings may yield similar performances, and Lloyd Morgan’s cannon warrants acceptance of the simpler explanation. Alternatively, when the different mechanisms interact with environmental conditions, variation in performance across environments might allow to statistically infer the mechanism responsible. However, to do this it is necessary to have quantitative predictions for the different candidate mechanisms. Here, we use a set of computational models to get quantitative predictions from alternative learning mechanisms, and using Bayesian statistics fit the model parameters to performance data. We used experimental data on performance in an ephemeral reward task by wild-caught cleaner fish *Labroides dimidiatus*, as well as cleaner and client fish densities from the locations of capture. The task can in principle be solved by estimating future consequences of an action, or by perceiving the removal of the ephemeral reward as psychological punishment (negative reinforcement). We found that a model where cleaners only estimate the future consequences of their actions explains best which cleaner relative abundances cause the fish to develop a preference for an ephemeral food source. This model also yields performances that can be considered the result of locally optimal decision-rules, in contrast to the negative reinforcement model.  We argue that the combination of computational models with data is a powerful tool to infer the mechanistic underpinning behind cognitive performance. 
keywords: learning, behaviour, cleaners, bayesian statisitics
---
